---
title: "IOS BEN models"
author: "Owen Exeter"
date: "20/06/2024"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#### Libraries
```{r results='hide', message=FALSE, warning=FALSE,echo=FALSE, include=FALSE}

library(tidyverse)
library(vegan)
library(MASS)
library(plotrix)  
library(grid)
library(gridExtra)
library(scales)
library(arm)
library(ggsignif)
library(ggpubr)
library(rstatix)
library(multcompView)
library(ggthemes)
library(data.table)
library(spatialreg)
library(spatialEco)
library(spdep)
library(sf)
library(dplyr)
library(ggplotify)
library(MuMIn)


```

### 0 | Read in ecological and metadata
```{r, echo=FALSE}
### Data processing
## read in BRUVS observation data
species_obs <- read.csv('~/IOS_benchmark_BRUVS_2025/EA_supporting_code/Data/IoS_BEN_observations.csv')
## just demersal species
reef_fish <- read.csv('~/IOS_benchmark_BRUVS_2025/EA_supporting_code/Data/IoS_BEN_observations_demerselfish.csv')
## read in lengths data
species_lengths <- read.csv('~/IOS_benchmark_BRUVS_2025/EA_supporting_code/Data/IoS_BEN_Lengths.csv')

## read in processed biomass data
site_biomass <- read.csv('~/IOS_benchmark_BRUVS_2025/EA_supporting_code/Data/IOS_BEN_biomass.csv')
## read in metadata
envo_variables <- read.csv('~/IOS_benchmark_BRUVS_2025/EA_supporting_code/Data/IoS_BEN_metadata.csv')

```

### 1 | BEN BRUVS SURVEY 
#### 1.1 Read in and calculate ecological indicies from BRUV survey data
```{r}

## calculate species richness
deployment_richness <- reef_fish %>%
  group_by(GRTS_ID) %>%
  summarise(Richness = length(Species))
## calculate mean maxN
deployment_maxn <- reef_fish %>%
  group_by(GRTS_ID) %>%
  summarise(Mean_MaxN = mean(MaxN))
## merge
mean_values1 <- left_join(deployment_maxn,deployment_richness, by='GRTS_ID')

## calculate total maxN
deployment_abundance <- reef_fish %>%
  group_by(GRTS_ID) %>%
  summarise(Total_MaxN = sum(MaxN))
deployment_abundance$Total_MaxN <-deployment_abundance$Total_MaxN
## merge
mean_values2 <- left_join(mean_values1,deployment_abundance, by='GRTS_ID')

## calculate shannons divserity 
div <- species_obs[c('GRTS_ID','Species','MaxN')] 
shannon_diversity <- div %>%
  group_by(GRTS_ID) %>%
  summarise(shannon_index = diversity(MaxN))
shannon_diversity_site <- shannon_diversity[c('GRTS_ID','shannon_index')]
## merge
mean_values3 <-  left_join(mean_values2,shannon_diversity_site, by='GRTS_ID')

## calculate mean length per sample 
## subset to just benthic associated species 
benthic_list <- as.list(unique(reef_fish$Species))
species_lengths <- subset(species_lengths,species_lengths$Species %in% benthic_list)
## calculate mean length
deployment_length <- species_lengths %>%
  group_by(GRTS_ID) %>%
  summarise(Mean_length = mean(Length))
## convert to cm
deployment_length$Mean_length <- deployment_length$Mean_length/10
## merge
mean_values4 <-  left_join(mean_values3,deployment_length, by='GRTS_ID')

## calculate total biomass per sample 
site_biomass <- site_biomass[c('GRTS_ID','Total_biomass_kg')]
mean_values5 <- left_join(mean_values4,site_biomass, by='GRTS_ID')

## join environmental variables of interest 
## remove duplicate rows
envo_variables <- envo_variables[,2:27]
envo_variables <- distinct(envo_variables)

## subset columns of interest
envo_variables_short <- envo_variables[,c('GRTS_ID','Date','Broadscale_habitat','Habitat','Mixed','Relief_score','Lat','Lon','TimeIN','Depth','Temp','Rug_range','Remoteness','Exposure','Slope_mean','MCZ','SAC','Height_of_tide')]
## Re-define habitat and relief as a categorical variables
envo_variables_short$Habitat <- as.factor(envo_variables_short$Habitat)      
envo_variables_short$Broadscale_habitat <- as.factor(envo_variables_short$Broadscale_habitat)      
envo_variables_short$Relief_score <- as.factor(envo_variables_short$Relief_score)      
## merge
IOS_data <- merge(mean_values5,envo_variables_short, by='GRTS_ID')

## assign a year and month
IOS_data$Year <- substr(IOS_data$Date,7,10)
envo_variables_short$Year <- substr(envo_variables_short$Date,7,10)
IOS_data$Month <- substr(IOS_data$Date,4,5)

## remove SVnStns for later analysis
IOS_data_SAC <- subset(IOS_data,IOS_data$SAC > 0)

```


#### 1.2 Explore the data and stats for text
```{r, results='hide'}
## Richness
mean(IOS_data$Richness)
var(IOS_data$Richness)
sd(IOS_data$Richness)
hist(IOS_data$Richness)
dispersion_index_rich <- var(IOS_data$Richness) / mean(IOS_data$Richness)
dispersion_index_rich

## shannon_index
mean(IOS_data$shannon_index)
var(IOS_data$shannon_index)
sd(IOS_data$shannon_index)
hist(IOS_data$shannon_index)

## Total maxn
mean(IOS_data$Total_MaxN)
var(IOS_data$Total_MaxN)
sd(IOS_data$Total_MaxN)
hist(IOS_data$Total_MaxN)
dispersion_index_maxn <- var(IOS_data$Total_MaxN) / mean(IOS_data$Total_MaxN)
dispersion_index_maxn

## Length
IOS_data_length <- na.omit(IOS_data)
IOS_data_length$Mean_length <- as.numeric(IOS_data_length$Mean_length)
mean(IOS_data_length$Mean_length)
var(IOS_data_length$Mean_length)
hist(IOS_data_length$Mean_length)

## Biomass
mean(IOS_data$Total_biomass_kg)
var(IOS_data$Total_biomass_kg)
sd(IOS_data$Total_biomass_kg)
hist(IOS_data$Total_biomass_kg)


```

#### 1.3 Proportion plots for Figure 2
```{r, warning=FALSE}
## read in obs data with 0 values for non present sites
species <- read.csv('~/IOS_benchmark_BRUVS_2025/EA_supporting_code/Data/iNEXT_data_format2.csv')
## convert wide to long
long <- melt(setDT(species), id.vars = "ID", variable.name = "Species")

## rename columns
colnames(long) <- c('GRTS_ID','Species','MaxN')
long$GRTS_ID <- gsub("IOS_","",as.character(long$GRTS_ID))
long$GRTS_ID <- as.integer(long$GRTS_ID)

#species mean maxn
maxN_mean <- long %>%
  group_by(Species) %>%
  summarise(Mean_MaxN = mean(MaxN))
maxN_sd <- long %>%
  group_by(Species) %>%
  summarise(SD = sd(MaxN))
maxN_sum <- long %>%
  group_by(Species) %>%
  summarise(Total_MaxN = sum(MaxN))
species_maxn<-cbind(maxN_mean,maxN_sd,maxN_sum)
species_maxn<-species_maxn[,c(1,2,4,6)]

## bind the families to species using merged Obs and meta data
family_habs <- read.csv('~/IOS_benchmark_BRUVS_2025/EA_supporting_code/Data/IoS_BEN_metadata_obs.csv')
habitat <- family_habs[,c('GRTS_ID','Broadscale_habitat')]
habitat$Habitat <- as.factor(ifelse(habitat$Broadscale_habitat == 'Circalittoral rock' | habitat$Broadscale_habitat ==  'Infralittoral rock','Reefs','Sediment'))
habitat<-unique(habitat)
family <- family_habs[,c('Species','Species_latin','Family','Simplified_functional_group')]
family <- unique(family)

species_summary<-left_join(species_maxn,family,by=c('Species'))
species_summary<-unique(species_summary)

## bind habitats and families to long format observations
species_obs<-left_join(long,habitat,by=c('GRTS_ID'))
species_obs<-left_join(species_obs,family,by=c('Species'))
species_obs <- unique(species_obs)

## number of positive occurrences by family per habitat type
family_hab <- subset(species_obs,species_obs$MaxN>0)
family_hab <- family_hab[,c('GRTS_ID','Family','Habitat')]
family_hab <- family_hab[!duplicated(family_hab),]
family_hab$count <- 1
hab_family_count <- family_hab %>%
  group_by(Family,Habitat) %>%
  summarise(Count = sum(count)) 
hab_family_prop <- hab_family_count
hab_family_prop$Proportion <- (hab_family_prop$Count / 280)*100

## filter top 10 most observed (in terms of number of deployments) families and group the rest together as 'Other' for plotting
top_ten <- hab_family_prop %>%
  group_by(Family) %>%
  summarise(Total = sum(Proportion))
top_ten_family <- top_n(top_ten,11,Total)
top_ten_family_list <- as.list(top_ten_family$Family)
all <- hab_family_prop[hab_family_prop$Family %in% top_ten_family_list,] 
all <- all[!all$Family == 'Other',] 
## create factor levels for plotting
all$Family <- factor(all$Family, levels = c("Labridae","Gadidae", "Ammodytidae", "Scyliorhinidae","Cancridae","Echinidae","Carcinidae","Majidae","Asteriidae","Palinuridae"))
all$Habitat <- factor(all$Habitat, levels = rev(c('Reefs','Sediment')))

## plot proportion of deployments by habitat
fam_prop_plot <- ggplot(data=all,aes(x=Proportion,y=Family,fill = Habitat)) + 
  geom_bar(position="stack", stat="identity",col= 'black',linewidth=0.2,width=0.9, alpha = 0.8)+
  scale_y_discrete(limits=rev)+
  scale_fill_manual(values=c("orange","#077DAA"))+
  scale_x_continuous(limits = c(0,100),expand = c(0, 0)) +
  geom_hline(yintercept=7.5,linetype = "twodash") +
  xlab('Proportion (%) of deployments')+
  theme_bw()+
  theme(legend.position=c(0.75,.25),legend.text=element_text(size=30),legend.title=element_text(size=30,face="bold"),axis.text.x =element_blank(),axis.text.y = element_text(size=20,face = "italic"),axis.title.y = element_blank(),axis.title.x = element_blank())+
  guides(fill=guide_legend(title="Habitat"),plot.margin = unit(c(0.5, 5, 0.5, 5), "mm"))
fam_prop_plot

#### boxplot of mean maxn for same families
## subset observations to top ten list 
maxn_family <- species_obs[,c('GRTS_ID','Family','MaxN','Habitat')]
maxn_family <- maxn_family[maxn_family$Family %in% top_ten_family_list,] 
maxn_family <- maxn_family[!maxn_family$Family == 'Other',]

## mean maxn per family and deployment
maxN_family_count <- maxn_family %>%
  group_by(Family,GRTS_ID) %>%
  summarise(Sum = sum(MaxN))
## mean + sd maxn by families and species
maxN_family_mean <- maxN_family_count %>%
  group_by(Family) %>%
  summarise(Mean_MaxN = mean(Sum))
maxN_family_sd <- maxN_family_count %>%
  group_by(Family) %>%
  summarise(SD = sd(Sum))

##add in habitat for point colours
habs<-unique(species_obs[,c('GRTS_ID','Habitat')])
maxN_family_count<- maxN_family_count %>%
  left_join(habs, by = "GRTS_ID")
maxN_family_count$Habitat <- factor(maxN_family_count$Habitat, levels = rev(c('Reefs','Sediment')))
## order for plotting
maxN_family_count$Family <- factor(maxN_family_count$Family, levels = c("Labridae","Gadidae", "Ammodytidae" ,"Scyliorhinidae","Cancridae","Echinidae","Carcinidae","Majidae","Asteriidae","Palinuridae"))

## boxplot with point colours being habitat type
fam_maxn_plot <-ggplot() + 
  geom_point(data=maxN_family_count,aes(x = Sum, y = Family,fill=Habitat),size = 2, alpha = 0.6, shape = 21, position=position_jitter(0.1), pch=21, show.legend = F) +
  scale_fill_manual(values=c("orange","#077DAA"))+
  geom_boxplot(data=maxN_family_count,aes(x = Sum, y = Family),outlier.shape = NA, alpha = 0.6,varwidth = TRUE,fill='grey') + 
  scale_x_continuous(limits = c(0,15),expand = c(0, 0)) +
  scale_y_discrete(limits=rev)+
  geom_hline(yintercept=7.5,linetype = "twodash") +
  theme_bw() +
  #xlab('Mean MaxN')+
  theme(axis.text.x =  element_blank(),axis.title.y = element_blank(),axis.title.x = element_blank(),axis.text.y = element_blank(),plot.margin = unit(c(0.5, 5, 0.5, 5), "mm"))
fam_maxn_plot

#### repeat process for species
## number of positive occurrences by family per habitat type
species_hab <- species_obs[,c('GRTS_ID','Species_latin','Habitat','MaxN')]
species_hab <- species_hab[!duplicated(species_hab),]
species_hab <- subset(species_hab,species_hab$MaxN>0)
species_hab$count <- 1
hab_species_count <- species_hab %>%
  group_by(Species_latin,Habitat) %>%
  summarise(Count = sum(count)) 

## number of positive occurrences by family per habitat type
hab_species_prop <- hab_species_count
hab_species_prop$Proportion <- (hab_species_prop$Count / 280)*100

## sum hab values for overall proportion
species_prop_total <- hab_species_prop %>%
  group_by(Species_latin) %>%
  summarise(Count = sum(Proportion)) 

## filter top 10 most observed (in terms of number of deployments) families and group the rest together as 'Other'
top_ten <- hab_species_prop %>%
  group_by(Species_latin) %>%
  summarise(Total = sum(Proportion))
top_ten_species <- top_n(top_ten,11,Total)
top_ten_species_list <- as.list(top_ten_species$Species_latin)
all2 <- hab_species_prop[hab_species_prop$Species_latin %in% top_ten_species_list,] 
all2 <- na.omit(all2)

## factor levels for plotting
all2$Species_latin <- factor(all2$Species_latin, levels = c("Pollachius pollachius","Labrus bergylta", "Labrus mixtus", "Ctenolabrus rupestris","Centrolabrus exoletus","Symphodus melops","Trisopterus luscus","Scyliorhinus canicula","Cancer pagurus","Echinus esculentus"))
all2$Species_latin[is.na(all2$Species_latin)] <- "Echinus esculentus"
all2$Habitat <- factor(all2$Habitat, levels = rev(c('Reefs','Sediment')))

## plot proportion of deployments by habitat
spp_prop_plot <- ggplot(data=all2,aes(x=Proportion,y=Species_latin,fill = Habitat)) + 
  geom_bar(position="stack", stat="identity",col= 'black',linewidth=0.2,width=0.9, alpha = 0.8)+
  scale_y_discrete(limits=rev)+
  scale_fill_manual(values=c("orange","#077DAA"))+
  scale_x_continuous(limits = c(0,100),expand = c(0, 0)) +
  geom_hline(yintercept=3.5,linetype = "twodash") +
  xlab('Proportion (%) of deployments')+
  theme_bw()+
  theme(legend.position = "none",axis.text.x = element_text(size=16),axis.text.y = element_text(size=20,face = "italic"),axis.title.y = element_blank(),axis.title.x =element_text(size=20),plot.margin = unit(c(0.5, 5, 0.5, 5), "mm"))
spp_prop_plot 

#### boxplot of mean maxn for same species
## subset observations to top ten list 
maxn_species <- species_obs[,c('GRTS_ID','Species_latin','MaxN')]
maxn_species <- maxn_species[maxn_species$Species_latin %in% top_ten_species_list,] 

## maxn per species and deployment
maxN_species_count <- maxn_species %>%
  group_by(Species_latin,GRTS_ID) %>%
  summarise(Mean_MaxN = mean(MaxN))

## mean maxn/sd per species
mean_maxN_species <- maxn_species %>%
  group_by(Species_latin) %>%
  summarise(Mean_MaxN = mean(MaxN))
sd_maxN_species <- maxn_species %>%
  group_by(Species_latin) %>%
  summarise(SD = sd(MaxN))

## exclude sandeels due to hyperabundance
maxN_species_count <- na.omit(maxN_species_count)

## order for plotting
maxN_species_count$Species_latin <- factor(maxN_species_count$Species_latin, levels = c("Pollachius pollachius","Labrus bergylta", "Labrus mixtus", "Ctenolabrus rupestris","Centrolabrus exoletus","Symphodus melops","Trisopterus luscus","Scyliorhinus canicula","Cancer pagurus","Echinus esculentus"))
maxN_species_count<- maxN_species_count %>%
  left_join(habs, by = "GRTS_ID")
maxN_species_count$Habitat <- factor(maxN_species_count$Habitat, levels = rev(c('Reefs','Sediment')))

## boxplot with point colours being habitat type
spp_maxn_plot <-ggplot() + 
  geom_point(size = 2, alpha = 0.4, shape = 21,position=position_jitter(0.1), pch=21, show.legend = F) +
  geom_boxplot(outlier.shape = NA, alpha = 0.9,varwidth = TRUE,fill='grey') + 
  #scale_fill_manual(values = c('#CD534CFF','#EFC000FF','#7AA6DCFF','#003C67FF'))+
  geom_point(data=maxN_species_count,aes(x = Mean_MaxN, y = Species_latin,fill=Habitat),size = 2, alpha = 0.6, shape = 21, position=position_jitter(0.1), pch=21, show.legend = F) +
  scale_fill_manual(values=c("orange","#077DAA"))+
  geom_boxplot(data=maxN_species_count,aes(x = Mean_MaxN, y = Species_latin),outlier.shape = NA, alpha = 0.6,varwidth = TRUE,fill='grey') + 
  scale_x_continuous(limits = c(0,35),expand = c(0, 0)) +
  scale_y_discrete(limits=rev)+
  geom_hline(yintercept=3.5,linetype = "twodash") +
  theme_bw() +
  xlab('Relative adundance (MaxN)')+
  theme(axis.text.x = element_text(size=16),axis.title.x =element_text(size=20),axis.text.y = element_blank(),axis.title.y = element_blank(),plot.margin = unit(c(0.5, 5, 0.5, 5), "mm"))
spp_maxn_plot

#### combine the plots for Figure 2
pp1 <- fam_prop_plot  /  spp_prop_plot
pp2 <- fam_maxn_plot / spp_maxn_plot
Figure2 <- pp1 | pp2 
Figure2



```

#### 1.4 Length histogram plots for Figure 3
```{r, warning=FALSE}

## select list of key indicator/commercially targeted species from list of data suitable species
indicator_spp <- as.list(c("Small_spotted_catshark","Nursehound","Pollack","Ballan_wrasse","Cuckoo_wrasse","Edible_crab","Common_spiny_lobster"))
indicator_lngths <- subset(species_lengths,species_lengths$Species %in% indicator_spp)

#### histogram plot of length distributions

## set order of species for logical plotting
indicator_lngths$Species_latin <-  factor(indicator_lngths$Species_latin, levels=c("Scyliorhinus canicula","Scyliorhinus stellaris","Pollachius pollachius","Labrus bergylta","Labrus mixtus","Cancer pagurus","Palinurus elephas"))

## generate lines of maturity lengths from Fishbase if available (exclude if not)
maturity <- data.frame(Species_latin = c("Scyliorhinus canicula","Scyliorhinus stellaris","Pollachius pollachius","Labrus bergylta","Labrus mixtus","Cancer pagurus","Palinurus elephas"), Z = c(500,780,410,170,160,NA,NA))
maturity$Species_latin <-  factor(maturity$Species_latin, levels=c("Scyliorhinus canicula","Scyliorhinus stellaris","Pollachius pollachius","Mullus surmuletus","Labrus bergylta","Labrus mixtus","Cancer pagurus","Palinurus elephas"))

## generate lines of MLS lengths from IOS IFCA if available (exclude if not)
size_limit <- data.frame(Species_latin = c("Scyliorhinus canicula","Scyliorhinus stellaris","Pollachius pollachius","Labrus bergylta","Labrus mixtus","Cancer pagurus","Palinurus elephas"), Z = c(NA,NA,300,NA,NA,140,110))
size_limit$Species_latin <-  factor(size_limit$Species_latin, levels=c("Scyliorhinus canicula","Scyliorhinus stellaris","Pollachius pollachius","Labrus bergylta","Labrus mixtus","Cancer pagurus","Palinurus elephas"))
## generate lines of MLS lengths from IOS IFCA if available (exclude if not)
size_limit2 <- data.frame(Species_latin = c("Scyliorhinus canicula","Scyliorhinus stellaris","Pollachius pollachius","Labrus bergylta","Labrus mixtus","Cancer pagurus","Palinurus elephas"), Z = c(NA,NA,NA,NA,NA,160,NA))
size_limit2$Species_latin <-  factor(size_limit2$Species_latin, levels=c("Scyliorhinus canicula","Scyliorhinus stellaris","Pollachius pollachius","Labrus bergylta","Labrus mixtus","Cancer pagurus","Palinurus elephas"))
indicator_lngths

## ggplot by without type
Figure3 <- ggplot(indicator_lngths, aes(x=Length,fill=Habitat))+
  geom_histogram(col='black',size=0.5,alpha=0.7)+
  scale_fill_manual(values=c("#077DAA","orange"),name = "Main habitat")+
  geom_vline(data = maturity, aes(xintercept = Z),linetype = "twodash",col='darkblue',linewidth=1.1)+
  geom_vline(data = size_limit, aes(xintercept = Z),linetype = "twodash",col='red',linewidth=1.1)+
  geom_vline(data = size_limit2, aes(xintercept = Z),linetype = "twodash",col='red',linewidth=1.1)+
  facet_wrap(Species_latin ~ ., scales="free", ncol = 2)+
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), breaks = ~round(unique(pretty(.))))+
  labs(x = "Length (mm)", y = "Frequency")+
  theme_minimal()+
  theme(legend.position="none",axis.title = element_text(size = 18),axis.text = element_text(size = 16),strip.text.x = element_text(size = 16),panel.background = element_rect(fill = 'white', colour = 'white'))
Figure3 


```


### 2 | BEN modelling
#### 2.1 Pair tests for habitat types and years exploration
#### 2.1.2 Richness ~ habitats Figure 5 + S2
```{r, warning=FALSE}
## Richness against habitats
## Mean and sd stats for text
richness_habitat_mean <- IOS_data %>%
  group_by(Habitat) %>%
  summarise(Mean_richness = mean(Richness))
richness_habitat_sd <- IOS_data %>%
  group_by(Habitat) %>%
  summarise(sd = sd(Richness))
## sub-habitat
richness_subhabitat_mean <- IOS_data %>%
  group_by(Broadscale_habitat) %>%
  summarise(Mean_richness = mean(Richness))
richness_subhabitat_sd <- IOS_data %>%
  group_by(Broadscale_habitat) %>%
  summarise(sd = sd(Richness))

## Mann-Whitney U test - indices against habitat (non-parametric pair comparison)
wxt_hab_rich<-wilcox.test(Richness ~ Habitat, data = IOS_data)
## plot
richness_hab_bp <- ggboxplot(IOS_data, x = "Habitat", y = "Richness", fill = "Habitat",  palette =  c("#077DAA","orange"), alpha = 0.7)+
  ylim(2,11.8)+
  xlab('Habitat')+
  ylab('Richness (n.)')+
  geom_signif(annotation="p< 0.0001", y_position=11.2, xmin=1, xmax=2, tip_length = 0,col='black',textsize=8)+
  theme_classic(base_size = 30)+
  theme(legend.position="none")
richness_hab_bp 

## Kruskal-Wallis H test between sub-feature habitat (Kruskal-Wallis test - Richness due to over dispersion)
kruskal.test<-kruskal.test(Richness ~ Broadscale_habitat, data = IOS_data)
## Pairwise comparison results
pairwise_rich_subhabs <- pairwise.wilcox.test(IOS_data$Richness, IOS_data$Broadscale_habitat,  p.adjust.method = "bonferroni")
z.score <- qnorm(pairwise_rich_subhabs$p.value/2)

## assign significance labels for plot
IOS_data$Cld='a'
labels <- IOS_data[,c('Broadscale_habitat','Cld')]
labels<- unique(labels)
labels$Cld<-c('b','c','a','bc')
pos <- IOS_data %>%
  group_by(Broadscale_habitat) %>% 
  summarise(mean=mean(Richness), sd=sd(Richness)) %>% 
  arrange(desc(mean))
lblpos<-merge(pos,labels)

## plot
## assign factor levels for habitat order
IOS_data$Broadscale_habitat<-factor(IOS_data$Broadscale_habitat, levels = c('Circalittoral rock','Infralittoral rock','Subtidal seagrass','Subtidal sediments'))
geom.text.size = 10

richness_subhab_bp <- ggboxplot(IOS_data, x = "Broadscale_habitat", y = "Richness", fill = "Broadscale_habitat",  palette =  c('#003C67FF','#7AA6DCFF','#CD534CFF','#EFC000FF'), alpha = 0.7)+
  labs(x = "Sub-habitat") +
  scale_x_discrete(labels = label_wrap(10)) +
  ylim(2,11.8)+
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position="none")+ 
  geom_text(data = lblpos, aes(label=Cld,x=Broadscale_habitat,y=11.5), size = 20/.pt)
richness_subhab_bp 

## merge for exporting
richness_hab_plot <- grid.arrange(richness_hab_bp, richness_subhab_bp , ncol=2)

#### results - 
richness_habitat_mean
richness_habitat_sd
richness_subhabitat_mean
richness_subhabitat_sd
wxt_hab_rich
pairwise_rich_subhabs
z.score 

```

#### 2.1.2 Total MaxN ~ habitats Figure 5 + S2
```{r, warning=FALSE}
## stats for text
maxn_habitat_sum <- IOS_data %>%
  group_by(Habitat) %>%
  summarise(Total_maxn = mean(Total_MaxN))
maxn_habitat_sd <- IOS_data %>%
  group_by(Habitat) %>%
  summarise(sd = sd(Total_MaxN))

## sub-habitat
maxn_subhabitat_mean <- IOS_data %>%
  group_by(Broadscale_habitat) %>%
  summarise(Total_MaxN = mean(Total_MaxN))
maxn_subhabitat_sd <- IOS_data %>%
  group_by(Broadscale_habitat) %>%
  summarise(sd = sd(Total_MaxN))

## Mann-Whitney U test - indices against habitat (non-parametric pair comparison)
wxt_hab_maxn<-wilcox.test(Total_MaxN ~ Habitat, data = IOS_data)

## plot
maxn_hab_bp <- ggboxplot(IOS_data, x = "Habitat", y = "Total_MaxN", fill = "Habitat",  palette =  c("#077DAA","orange"), alpha = 0.7)+
  ylab("Relative abundance (n.)")+
  ylim(0,200)+
  xlab('Habitat')+
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 14))+ 
  geom_signif(annotation="ns (p= 0.08)", y_position=190, xmin=1, xmax=2, tip_length = 0,col='black',textsize=8)+
  theme_classic(base_size = 30) +
  theme(legend.position="none")
maxn_hab_bp 

## Kruskal-Wallis H test between sub-feature habitat (Kruskal-Wallis test - Richness due to over dispersion)
kruskal.test<-kruskal.test(Total_MaxN ~ Broadscale_habitat, data = IOS_data)
## Print the pairwise comparison results
pairwise_maxn_subhabs <- pairwise.wilcox.test(IOS_data$Total_MaxN, IOS_data$Broadscale_habitat,  p.adjust.method = "bonferroni")

## assign significance labels for plot
# assign significance labels for plot
labels <- IOS_data[,c('Broadscale_habitat','Cld')]
labels<- unique(labels)
labels$Cld<-c('bc','bc','ac','abc')
pos <- IOS_data %>%
  group_by(Broadscale_habitat) %>% 
  summarise(total=mean(Total_MaxN), sd=sd(Total_MaxN)) %>% 
  arrange(desc(total))
lblpos<-merge(pos,labels)

maxn_subhab_bp <- ggboxplot(IOS_data, x = "Broadscale_habitat", y = "Total_MaxN", fill = "Broadscale_habitat",  palette =  c('#003C67FF','#7AA6DCFF','#CD534CFF','#EFC000FF'), alpha = 0.7)+
  labs(x = "Sub-habitat") +
  scale_x_discrete(labels = label_wrap(10)) +
  ylab("Relative abundance")+
  ylim(0,200)+
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position="none")+ 
  geom_text(data = lblpos, aes(label=Cld,x=Broadscale_habitat,y=190), size = 20/.pt)
maxn_subhab_bp 

## merge for exporting
maxn_hab_plot <- grid.arrange(maxn_hab_bp, maxn_subhab_bp , ncol=2)

#### results 
maxn_habitat_sum
maxn_habitat_sd
maxn_subhabitat_mean
maxn_subhabitat_sd
wxt_hab_maxn
pairwise_maxn_subhabs


```

#### 2.1.3 Total biomass ~ habitats Figure 5 + S2
```{r}
## stats for text
biom_habitat_mean <- IOS_data %>%
  group_by(Habitat) %>%
  summarise(Mean_biom = mean(Total_biomass_kg))
biom_habitat_sd <- IOS_data %>%
  group_by(Habitat) %>%
  summarise(sd = sd(Total_biomass_kg))

## sub-habitat
biom_subhabitat_mean <- IOS_data %>%
  group_by(Broadscale_habitat) %>%
  summarise(Mean_richness = mean(Total_biomass_kg))
biom_subhabitat_sd <- IOS_data %>%
  group_by(Broadscale_habitat) %>%
  summarise(sd = sd(Total_biomass_kg))

## Mann-Whitney U test - indices against habitat (non-parametric pair comparison)
wxt_hab_biom<-wilcox.test(Total_biomass_kg ~ Habitat, data = IOS_data)

## plot
biom_hab_bp <- ggboxplot(IOS_data, x = "Habitat", y = "Total_biomass_kg", fill = "Habitat",  palette =  c("#077DAA","orange"), alpha = 0.7)+
  ylab("Biomass (kg)")+
  xlab('Habitat')+
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 14))+ 
  geom_signif(annotation="p< 0.0001", y_position=190, xmin=1, xmax=2, tip_length = 0,col='black',textsize=8)+
  theme_classic(base_size = 30)+
  theme(legend.position="none")
biom_hab_bp 

## Kruskal-Wallis H test between sub-feature habitat (Kruskal-Wallis test - Richness due to over dispersion)
kruskal.test<-kruskal.test(Total_biomass_kg ~ Broadscale_habitat, data = IOS_data)
## Print the pairwise comparison results
pairwise_biom_subhabs <- pairwise.wilcox.test(IOS_data$Total_biomass_kg, IOS_data$Broadscale_habitat,  p.adjust.method = "bonferroni")

## assign significance labels for plot
labels2 <- IOS_data[,c('Broadscale_habitat','Cld')]
labels2<- unique(labels2)
labels$Cld<-c('b','bc','a','c')
pos <- IOS_data %>%
  group_by(Broadscale_habitat) %>% 
  summarise(mean=mean(Total_biomass_kg), sd=sd(Total_biomass_kg)) %>% 
  arrange(desc(mean))
lblpos<-merge(pos,labels)

biom_subhab_bp <- ggboxplot(IOS_data, x = "Broadscale_habitat", y = "Total_biomass_kg", fill = "Broadscale_habitat",  palette =  c('#003C67FF','#7AA6DCFF','#CD534CFF','#EFC000FF'), alpha = 0.7)+
  labs(x = "Sub-habitat") +
  scale_x_discrete(labels = label_wrap(10)) +
  ylab("Total biomass (kg)")+
  #ylim(0,50)+
  theme(axis.text = element_text(size = 18), axis.title = element_text(size = 18), legend.position="none")+ 
  geom_text(data = lblpos, aes(label=Cld,x=Broadscale_habitat,y=200), size = 20/.pt)
biom_subhab_bp 

#### results -
biom_habitat_mean
biom_habitat_sd
biom_subhabitat_mean
biom_subhabitat_sd
wxt_hab_biom
pairwise_biom_subhabs

```

#### 2.1.4 All indices ~ Year Figure S2
```{r}
### check if year has a significant impact prior to further testing
## Mann-Whitney U test - indices against year 
wilcox.test(Richness ~ Year, data = IOS_data)
wilcox.test(Total_MaxN ~ Year, data = IOS_data)
wilcox.test(Total_biomass_kg ~ Year, data = IOS_data)
wilcox.test(Total_biomass_kg ~ Year, data = IOS_data_SAC)

## plot
bxpyr <- ggboxplot(IOS_data, x = "Year", y = "Richness", alpha = 0.7)+ 
  geom_signif(annotation="ns (p= 0.14)", y_position=11.1, xmin=1, xmax=2, tip_length = 0,col='black')+
  theme(axis.text = element_text(size = 12), axis.title = element_text(size = 14),legend.position="none")
bxpyr2 <- ggboxplot(IOS_data, x = "Year", y = "Total_MaxN", alpha = 0.7)+ 
  geom_signif(annotation="ns (p= 0.11)", y_position=405, xmin=1, xmax=2, tip_length = 0,col='black')+
  ylab('Total MaxN')+
  theme(axis.text = element_text(size = 12),  axis.title = element_text(size = 14),legend.position="none")
bxpyr3 <- ggboxplot(IOS_data, x = "Year", y = "Total_biomass_kg", alpha = 0.7)+ 
  scale_y_continuous(limits = c(0,210))+
  geom_signif(annotation="ns (p= 0.07)", y_position=205, xmin=1, xmax=2, tip_length = 0,col='black')+
  ylab('Biomass (kg) all')+
  theme(axis.text = element_text(size = 12),  axis.title = element_text(size = 14),legend.position="none")
bxpyr4 <- ggboxplot(IOS_data_SAC, x = "Year", y = "Total_biomass_kg", alpha = 0.7)+ 
  ylab('Biomass (kg) SAC')+ 
  geom_signif(annotation="ns (p= 0.18)", y_position=40, xmin=1, xmax=2, tip_length = 0,col='black')+
  theme(axis.text = element_text(size = 12),  axis.title = element_text(size = 14),legend.position="none")


```

#### 3.1 GLMs - Richness ~ predictor variables
```{r}
## Richness
# Create a spatial weights matrix
Moran_data <- IOS_data
coords <- as.matrix(Moran_data[, c("Lat", "Lon")]) # Coordinates
nb <- knn2nb(knearneigh(coords, k = 4)) # 4-nearest neighbors
listw <- nb2listw(nb, style = "W") # Spatial weights list
# Compute the spatially lagged response variable
IOS_data$W_y <- lag.listw(listw, IOS_data$Richness)
W_y <- lag.listw(listw, IOS_data$Richness)

## model parameters
null_model <- glm(Richness ~ 1 , data = IOS_data,  family=poisson(link=log))
model1 <- glm(Richness~Habitat + W_y + Year, data = IOS_data,  family=poisson(link=log))  
model2 <- glm(Richness~Depth +  W_y + Year, data = IOS_data,  family=poisson(link=log))  
model3 <- glm(Richness~Temp + W_y + Year, data = IOS_data,  family=poisson(link=log))    
model4 <- glm(Richness~Relief_score + W_y + Year,family=poisson(link=log),data = IOS_data)
model5 <- glm(Richness~Remoteness + W_y + Year, data = IOS_data,  family=poisson(link=log)) 
model6 <- glm(Richness~Exposure + W_y + Year, data = IOS_data,  family=poisson(link=log))
model7 <- glm(Richness~Depth*Habitat + W_y + Year, data = IOS_data,  family=poisson(link=log))   
model8 <- glm(Richness~Temp*Habitat + W_y + Year, data = IOS_data,  family=poisson(link=log))    
model9 <- glm(Richness~Relief_score*Habitat + W_y + Year,family=poisson(link=log),data = IOS_data)
model10 <- glm(Richness~Remoteness*Habitat + W_y + Year, data = IOS_data,  family=poisson(link=log))  
model11 <- glm(Richness~Exposure*Habitat + W_y + Year, data = IOS_data,  family=poisson(link=log)) 
model12 <- glm(Richness~Relief_score + Depth + W_y + Year, data = IOS_data,  family=poisson(link=log))
model13 <- glm(Richness~Relief_score + Remoteness + W_y + Year, data = IOS_data,  family=poisson(link=log))  
model14 <- glm(Richness~Relief_score + Exposure + W_y + Year, data = IOS_data,  family=poisson(link=log)) 
model15 <- glm(Richness~Remoteness + Exposure + W_y + Year, data = IOS_data,  family=poisson(link=log))  
model16 <- glm(Richness~Depth + Remoteness + W_y + Year,family=poisson(link=log),data = IOS_data)

## model selection table
(msAICc <- model.sel(null_model, model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
df<-as.data.frame(msAICc,col_names = T)
df$model <- row.names(df)
df$model <- as.numeric(substring(df$model, 6))
df$model<-as.character(df$model)
df <- df %>% mutate(model = replace_na(model,'null_model'))

## Extract information for specific models
summary(model12)
summary(model7)
##check plots
par(mfrow = c(2, 2))
plot(model12)
hist(model12$residuals)

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})
model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs <- left_join(model_outputs,pseudo_r_squared_df)

## calulate LRT
logLik_null <- logLik(null_model)
logLik_full <- logLik(model12)

# Likelihood ratio test statistic
LRT_stat <- 2 * (as.numeric(logLik_full) - as.numeric(logLik_null))

# Degrees of freedom (difference in the number of parameters)
df <- attr(logLik_full, "df") - attr(logLik_null, "df")

# p-value from the chi-squared distribution
p_value <- pchisq(LRT_stat, df = df, lower.tail = FALSE)

# Calculate Moran's I for each model
moran_results_richness <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  residuals <- residuals(model)
  moran_test <- moran.test(residuals, listw)
    data.frame(
    model = paste0(i-1),
    moran_I_richness = moran_test$estimate[1],
    sd_richness = moran_test$statistic,
    p_value_richness = moran_test$p.value             
  )
}))

## plot the top variables present in the optimal model
## generate modeled values for depth and relief against richness
## depth
## for reef
modelx <- glm(Richness~Depth*Habitat +W_y+ Year, data = IOS_data,  family=poisson(link=log))   
pdat <- expand.grid(Depth = seq(1.6,44,0.5),  Habitat = "Reefs", W_y=mean(IOS_data$W_y),Year=as.character(names(sort(table(IOS_data$Year), decreasing = TRUE)[1])))
pred <- predict (modelx, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)
predframe <- data.frame (pdat, Total_MaxN = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Total_MaxN - 1.96*predframe$se
predframe$upr <- predframe$Total_MaxN + 1.96*predframe$se
## for sediment
pdat2 <- expand.grid(Depth = seq(2.4,44,0.5),  Habitat = "Sediment",W_y=mean(IOS_data$W_y), Year=as.character(names(sort(table(IOS_data$Year), decreasing = TRUE)[1])))
pred2 <- predict (modelx, newdata = pdat2, na.rm = T, type= "response", se.fit = TRUE)
predframe2 <- data.frame (pdat2, Total_MaxN = pred2$fit, se = pred2$se.fit)
predframe2$lwr <- predframe2$Total_MaxN - 1.96*predframe2$se
predframe2$upr <- predframe2$Total_MaxN + 1.96*predframe2$se
## bind
predframe3 <- rbind(predframe,predframe2)

depth_richness_gg <- ggplot(mapping = aes(x = Depth)) +
  geom_line(aes(y = Total_MaxN, colour = Habitat), data = predframe3,linewidth=1) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = Habitat), data = predframe3, alpha = 0.2)+
  geom_rug(aes(colour = Habitat), data = IOS_data)+
  scale_fill_manual(values=c("#077DAA","orange"))+
  scale_color_manual(values = c("#077DAA", "orange")) +
  labs(x = 'Depth (m)')+
  labs(y = '')+
  scale_x_continuous(limits = c(1.5,45), expand = c(0, 0)) +
  scale_y_continuous(limits = c(2,10), breaks=c(4,6,8,10), expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(legend.position="none", axis.title.y=element_blank())
depth_richness_gg 

##relief
modely <- glm(Richness~Relief_score + Depth +W_y+ Year, data = IOS_data,  family=poisson(link=log))  
pdat <- expand.grid(Relief_score = levels(factor(IOS_data$Relief_score)), Depth = mean(IOS_data$Depth), W_y=mean(IOS_data$W_y), Year=as.character(names(sort(table(IOS_data$Year), decreasing = TRUE)[1])))
pred <- predict (modely, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)

# combine the predictions with the predictors, 
predframe <- data.frame (pdat, Richness = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Richness - 1.96*predframe$se
predframe$upr <- predframe$Richness + 1.96*predframe$se

# Add the predicted means
relief_richness_gg<-ggplot(predframe, aes(x=Relief_score, y=Richness)) + 
  geom_pointrange(aes(ymin=lwr, ymax=upr),size=1,linewidth = 1) +
  #scale_color_manual(values=c("#077DAA","orange"),guide="none")+
  #geom_signif(annotation="***", y_position=6.5, xmin=1, xmax=2, tip_length = 0,col='black')+
  ylab('')+
  xlab('Relief score')+
  scale_y_continuous(limits = c(3,7.5), breaks=c(4,5,6,7), expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(legend.position="none", axis.title.y=element_blank())
relief_richness_gg

richness_plots <-  grid.arrange(richness_hab_bp,depth_richness_gg, relief_richness_gg , ncol=3)


```

#### 3.2 GLMs - MaxN ~ predictor variables
```{r}
## total maxn
# Create a spatial weights matrix
Moran_data <- IOS_data
coords <- as.matrix(Moran_data[, c("Lat", "Lon")]) # Coordinates
nb <- knn2nb(knearneigh(coords, k = 4)) # 4-nearest neighbors
listw <- nb2listw(nb, style = "W") # Spatial weights list
# Compute the spatially lagged response variable
IOS_data$W_y <- lag.listw(listw, IOS_data$Total_MaxN)
W_y <- lag.listw(listw, IOS_data$Total_MaxN)

## model parameters
null_model <- glm.nb(Total_MaxN ~ 1 , data = IOS_data,link=log)
model1 <- glm.nb(Total_MaxN~Habitat + W_y + Year, data = IOS_data, link=log)  
model2 <- glm.nb(Total_MaxN~Depth +  W_y + Year, data = IOS_data, link=log)  
model3 <- glm.nb(Total_MaxN~Temp + W_y + Year, data = IOS_data,  link=log)    
model4 <- glm.nb(Total_MaxN~Relief_score + W_y + Year,data = IOS_data, link=log)
model5 <- glm.nb(Total_MaxN~Remoteness + W_y + Year,data = IOS_data, link=log) 
model6 <- glm.nb(Total_MaxN~Exposure + W_y + Year, data = IOS_data,  link=log)
model7 <- glm.nb(Total_MaxN~Depth*Habitat + W_y + Year, data = IOS_data,  link=log)   
model8 <- glm.nb(Total_MaxN~Temp*Habitat + W_y + Year, data = IOS_data,  link=log)    
model9 <- glm.nb(Total_MaxN~Relief_score*Habitat + W_y + Year,link=log,data = IOS_data)
model10 <- glm.nb(Total_MaxN~Remoteness*Habitat + W_y + Year, data = IOS_data,  link=log)  
model11 <- glm.nb(Total_MaxN~Exposure*Habitat + W_y + Year, data = IOS_data,  link=log) 
model12 <- glm.nb(Total_MaxN~Relief_score + Depth + W_y + Year, data = IOS_data,  link=log)    
model13 <- glm.nb(Total_MaxN~Relief_score + Remoteness + W_y + Year, data = IOS_data,  link=log)  
model14 <- glm.nb(Total_MaxN~Relief_score + Exposure + W_y + Year, data = IOS_data,  link=log) 
model15 <- glm.nb(Total_MaxN~Remoteness + Exposure + W_y + Year, data = IOS_data,  link=log)  
model16 <- glm.nb(Total_MaxN~Depth + Remoteness + W_y + Year,link=log,data = IOS_data)

## model selection table
(msAICc <- model.sel(null_model, model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
df<-as.data.frame(msAICc,col_names = T)
df$model <- row.names(df)
df$model <- as.numeric(substring(df$model, 6))
df$model<-as.character(df$model)
df <- df %>% mutate(model = replace_na(model,'null_model'))

## Extract information for specific models
summary(model7)
## AIC
AIC(model7)
##check plots
par(mfrow = c(2, 2))
plot(model7)

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs <- left_join(model_outputs,pseudo_r_squared_df)

# Calculate Moran's I for each model
moran_results_maxn <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  residuals <- residuals(model)
  moran_test <- moran.test(residuals, listw)
    data.frame(
    model = paste0(i-1),
    moran_I_maxn = moran_test$estimate[1],        
    sd_maxn = moran_test$statistic,        
    p_value_maxn = moran_test$p.value              
  )
}))

## create dataset to predict to
## for reef
modelxy <- glm.nb(Total_MaxN~Depth*Habitat +W_y +Year, data = IOS_data,  link=log)
pdat <- expand.grid(Depth = seq(1.6,44,0.5),  Habitat = "Reefs",W_y=mean(IOS_data$W_y), Year=as.character(names(sort(table(IOS_data$Year), decreasing = TRUE)[1])))
pred <- predict (modelxy, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)
predframe <- data.frame (pdat, Total_MaxN = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Total_MaxN - 1.96*predframe$se
predframe$upr <- predframe$Total_MaxN + 1.96*predframe$se
## for sediment
pdat2 <- expand.grid(Depth = seq(2.4,44,0.5),  Habitat = "Sediment", W_y=mean(IOS_data$W_y), Year=as.character(names(sort(table(IOS_data$Year), decreasing = TRUE)[1])))
pred2 <- predict (modelxy, newdata = pdat2, na.rm = T, type= "response", se.fit = TRUE)
predframe2 <- data.frame (pdat2, Total_MaxN = pred2$fit, se = pred2$se.fit)
predframe2$lwr <- predframe2$Total_MaxN - 1.96*predframe2$se
predframe2$upr <- predframe2$Total_MaxN + 1.96*predframe2$se
## bind
predframe3 <- rbind(predframe,predframe2)

depthhab_totalmaxn_gg <- ggplot(mapping = aes(x = Depth)) +
  geom_line(aes(y = Total_MaxN, colour = Habitat), data = predframe3,linewidth=1) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = Habitat), data = predframe3, alpha = 0.2)+
  geom_rug(aes(colour = Habitat), data = IOS_data)+
  scale_fill_manual(values=c("#077DAA","orange"))+
  scale_color_manual(values = c("#077DAA", "orange")) +
  labs(x = 'Depth (m)')+
  labs(y = '')+
  scale_x_continuous(limits = c(1.5,45), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,200), breaks=c(50,100,150, 200), expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(legend.position="none", axis.title.y=element_blank())
depthhab_totalmaxn_gg

maxn_plots <-  grid.arrange(maxn_hab_bp,depthhab_totalmaxn_gg,ncol=3)

```

#### 3.2 GLMs - Biomass ~ predictor variables
```{r}
## biomass
# Create a spatial weights matrix
Moran_data <- IOS_data_SAC
coords <- as.matrix(Moran_data[, c("Lat", "Lon")]) # Coordinates
nb <- knn2nb(knearneigh(coords, k = 4)) # 4-nearest neighbors
listw <- nb2listw(nb, style = "W") # Spatial weights list
# Compute the spatially lagged response variable
IOS_data_SAC$W_y <- lag.listw(listw, IOS_data_SAC$Total_biomass_kg)
W_y <- lag.listw(listw, IOS_data_SAC$Total_biomass_kg)

## model parameters
null_model <- glm(Total_biomass_kg ~ 1, data = IOS_data_SAC, family = Gamma(link =log))
model1 <- glm(Total_biomass_kg~Habitat + W_y + Year, data = IOS_data_SAC, family = Gamma(link =log))  
model2 <- glm(Total_biomass_kg~Depth +  W_y + Year, data = IOS_data_SAC, family = Gamma(link =log))  
model3 <- glm(Total_biomass_kg~Temp + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log))    
model4 <- glm(Total_biomass_kg~Relief_score + W_y + Year,data = IOS_data_SAC, family = Gamma(link =log))
model5 <- glm(Total_biomass_kg~Remoteness + W_y + Year,data = IOS_data_SAC, family = Gamma(link =log)) 
model6 <- glm(Total_biomass_kg~Exposure + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log))
model7 <- glm(Total_biomass_kg~Depth*Habitat + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log))   
model8 <- glm(Total_biomass_kg~Temp*Habitat + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log))    
model9 <- glm(Total_biomass_kg~Relief_score*Habitat + W_y + Year,family = Gamma(link =log),data = IOS_data_SAC)
model10 <- glm(Total_biomass_kg~Remoteness*Habitat + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log))  
model11 <- glm(Total_biomass_kg~Exposure*Habitat + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log)) 
model12 <- glm(Total_biomass_kg~Relief_score + Depth + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log))    
model13 <- glm(Total_biomass_kg~Relief_score + Remoteness + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log))  
model14 <- glm(Total_biomass_kg~Relief_score + Exposure + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log)) 
model15 <- glm(Total_biomass_kg~Remoteness + Exposure + W_y + Year, data = IOS_data_SAC,  family = Gamma(link =log))  
model16 <- glm(Total_biomass_kg~Depth + Remoteness + W_y + Year,family = Gamma(link =log),data = IOS_data_SAC)

## model selection table
(msAICc <- model.sel(null_model, model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
df<-as.data.frame(msAICc,col_names = T)
df$model <- row.names(df)
df$model <- as.numeric(substring(df$model, 6))
df$model<-as.character(df$model)
df <- df %>% mutate(model = replace_na(model,'null_model'))

## Extract information for specific models
summary(model2)
## AIC
AIC(model2)
##check plots
par(mfrow = c(2, 2))
plot(model2)
## Extract information for specific models
summary(model16)
## AIC
AIC(model16)
##check plots
par(mfrow = c(2, 2))
plot(model16)
## Extract information for specific models
summary(model7)
## AIC
AIC(model7)
##check plots
par(mfrow = c(2, 2))
plot(model7)

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs <- left_join(model_outputs,pseudo_r_squared_df)

# Calculate Moran's I for each model
moran_results_biomass <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  residuals <- residuals(model)
  moran_test <- moran.test(residuals, listw)
    data.frame(
    model = paste0(i-1),
    moran_I_biomass = moran_test$estimate[1],   
    sd_biomass = moran_test$statistic,
    p_value_biomass = moran_test$p.value              
  )
}))

## merge all 3 Moran's I data frames and export for Supp. Table 
morans_outputs<-merge(moran_results_richness,moran_results_maxn) %>%
              merge(moran_results_biomass)
morans_outputs$model<-as.numeric(morans_outputs$model)


## plot the top variables present in the optimal model
## generate modeled values for depth and relief against richness
##depth
## create dataset to predict to
## for reef
modelx <- glm(Total_biomass_kg~Depth + W_y +Year, data = IOS_data, family = Gamma(link =log))  
pdat <- expand.grid(Depth = seq(1.6,44,0.5), W_y=mean(IOS_data_SAC$W_y),Year=as.character(names(sort(table(IOS_data_SAC$Year), decreasing = TRUE)[1])))
pred <- predict (modelx, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)
predframe <- data.frame (pdat, Total_biomass_kg = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Total_biomass_kg - 1.96*predframe$se
predframe$upr <- predframe$Total_biomass_kg + 1.96*predframe$se

depth_biomass_gg <- ggplot(mapping = aes(x = Depth)) +
  geom_line(aes(y = Total_biomass_kg), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  geom_rug(aes(), data = IOS_data_SAC)+
  labs(x = 'Depth (m)')+
  labs(y = '')+
  scale_x_continuous(limits = c(1.5,45), expand = c(0, 0)) +
  #scale_y_continuous(limits = c(3,12), breaks=c(3,5, 7, 9,11), expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(axis.title.y=element_blank())
depth_biomass_gg


## Remoteness
modely <-  glm(Total_biomass_kg~Remoteness+Depth+ W_y+Year,family = Gamma(link =log),data = IOS_data_SAC)
pdat <- expand.grid(Remoteness= seq(32.1,7048.7,10),Depth=mean(IOS_data_SAC$Depth), W_y=mean(IOS_data_SAC$W_y),Year=as.character(names(sort(table(IOS_data$Year), decreasing = TRUE)[1])))
pred <- predict (modely, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)
# combine the predictions with the predictors, 
predframe <- data.frame (pdat, Total_biomass_kg = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Total_biomass_kg - 1.96*predframe$se
predframe$upr <- predframe$Total_biomass_kg + 1.96*predframe$se

# Add the predicted means
remote_biomasss_gg<-ggplot(predframe, aes(x=Remoteness, y=Total_biomass_kg)) + 
  geom_line(aes(y = Total_biomass_kg), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  geom_rug(aes(), data = IOS_data_SAC,sides="b")+
  ylab('')+
  xlab('Remoteness (m)')+
  scale_x_continuous(limits = c(30,7050), expand = c(0, 0)) +
  scale_y_continuous(limits = c(2,10), breaks=c(4,6, 8,10),expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(legend.position="none", axis.title.y=element_blank())
remote_biomasss_gg

biomass_plots <-  grid.arrange(biom_hab_bp,depth_biomass_gg, remote_biomasss_gg , ncol=3)


```

## combine all GLM plots into matrix figure of model outputs of interest for Figure 5
```{r}
## combine all GLM plots into matrix figure of model outputs of interest
all_model_plots <-  grid.arrange(richness_plots,maxn_plots, biomass_plots , nrow=3)

```


#### 3.3 Species lengths modelling
### 3.3.1 Small spotted catshark 
```{r}
## small spotted catshark
indicator_spp <- species_lengths
ssc <- subset(indicator_spp,indicator_spp$Species == 'Small_spotted_catshark')
ssc <- ssc %>%
  group_by(GRTS_ID) %>%
  summarise(Length = mean(Length))
ssc <- merge(ssc,envo_variables_short, by=c('GRTS_ID'))
hist(ssc$Length)
mean_val<-mean(ssc$Length)
var(ssc$Length)
std.error(ssc$Length)    

## model parameters
# Fit the null model (intercept only) using fitme
null_model <- glm(Length ~ 1, data = ssc,  family="gaussian")
model1 <- glm(Length~Habitat + Year, data = ssc,  family="gaussian") 
model2 <- glm(Length~Depth  + Year, data = ssc,  family="gaussian")  
model3 <- glm(Length~Temp  + Year, data = ssc,  family="gaussian")    
model4 <- glm(Length~Relief_score  + Year,data = ssc, family="gaussian")
model5 <- glm(Length~Remoteness  + Year, data = ssc,  family="gaussian") 
model6 <- glm(Length~Exposure  + Year, data = ssc,  family="gaussian")
model7 <- glm(Length~Depth*Habitat  + Year, data = ssc,  family="gaussian")   
model8 <- glm(Length~Temp*Habitat  + Year, data = ssc,  family="gaussian")    
model9 <- glm(Length~Relief_score*Habitat  + Year,data = ssc ,family="gaussian")
model10 <- glm(Length~Remoteness*Habitat  + Year, data = ssc,  family="gaussian")  
model11 <- glm(Length~Exposure*Habitat  + Year, data = ssc,  family="gaussian") 
model12 <- glm(Length~Relief_score + Depth  + Year, data = ssc,  family="gaussian")    
model13 <- glm(Length~Relief_score + Remoteness  + Year, data = ssc,  family="gaussian")  
model14 <- glm(Length~Relief_score + Exposure  + Year, data = ssc,  family="gaussian") 
model15 <- glm(Length~Remoteness + Exposure  + Year, data = ssc,  family="gaussian")  
model16 <- glm(Length~Depth + Remoteness  + Year,data = ssc ,family="gaussian")

## model selection table
(msAICc <- model.sel(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
ssc_df<-as.data.frame(msAICc,col_names = T)
ssc_df$model <- row.names(ssc_df)
ssc_df$model <- as.numeric(substring(ssc_df$model, 6))
ssc_df$model<-as.character(ssc_df$model)
ssc_df$Species <- 'Small spotted catshark'
ssc_df <- ssc_df %>% mutate(model = replace_na(model,'null_model'))

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(ssc_df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs_ssc <- left_join(model_outputs,pseudo_r_squared_df)

## create dataset to predict to
modely <- glm(Length~Exposure + Year, data = ssc,  family="gaussian")
pdat <- expand.grid(Exposure= seq(6.4,175.6,1), Year=as.character(names(sort(table(ssc$Year), decreasing = TRUE)[1])))
pred <- predict (modely, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)

# Step 3: combine the predictions with the predictors, 
#         into a final dataframe (predframe)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se

# Add the predicted means
ssc_p<-ggplot(predframe, aes(x=Exposure, y=Length)) + 
  geom_line(aes(y = Length), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  coord_cartesian(xlim = c(min(predframe$Exposure), max(predframe$Exposure)), clip = "off")+
  geom_rug(aes(), data = ssc,sides="b")+
  ylab('')+
  xlab('Exposure ()')+
  scale_x_continuous(limits = c(6.4,180), expand = c(0, 0)) +
  scale_y_continuous(limits = c(550,700),expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(axis.title.y=element_blank())
  
ssc_p

## for reef
modely <- glm(Length~Exposure*Habitat , data = ssc,  family="gaussian")
pdat <- expand.grid(Exposure= seq(6.4,168.9,1),  Habitat = "Reefs")
pred <- predict (modely, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se
## for sediment
pdat2 <- expand.grid(Exposure = seq(13.4,175.6,1),  Habitat = "Sediment")
pred2 <- predict (modely, newdata = pdat2, na.rm = T, type= "response", se.fit = TRUE)
predframe2 <- data.frame (pdat2, Length = pred2$fit, se = pred2$se.fit)
predframe2$lwr <- predframe2$Length - 1.96*predframe2$se
predframe2$upr <- predframe2$Length + 1.96*predframe2$se
## bind
predframe3 <- rbind(predframe,predframe2)

ssc_p2 <- ggplot(mapping = aes(x = Exposure)) +
  geom_line(aes(y = Length, colour = Habitat), data = predframe3,linewidth=1) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = Habitat), data = predframe3, alpha = 0.2)+
  geom_rug(aes(colour = Habitat), data = ssc)+
  scale_fill_manual(values=c("#077DAA","orange"))+
  scale_color_manual(values = c("#077DAA", "orange")) +
  coord_cartesian(xlim = c(min(predframe$Exposure), max(predframe$Exposure)), clip = "off")+
  labs(x = 'Exposure ()')+
  labs(y = '')+
  scale_x_continuous(limits = c(0,176), expand = c(0, 0)) +
  scale_y_continuous(limits = c(500,750), expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(legend.position="none", axis.title.y=element_blank())
  
ssc_p2

```


### 3.3.2 Nursehound
```{r}

## Nursehound
nh <- subset(indicator_spp,indicator_spp$Species == 'Nursehound')
nh <- nh %>%
  group_by(GRTS_ID) %>%
  summarise(Length = mean(Length))
nh <- merge(nh,envo_variables_short, by='GRTS_ID')
 
hist(nh$Length)
mean(nh$Length)
var(nh$Length)
std.error(nh$Length) 

## model parameters
# Fit the null model (intercept only) using fitme
null_model <- glm(Length ~ 1, data = nh,  family="gaussian")
model1 <- glm(Length~Habitat + Year, data = nh,  family="gaussian") 
model2 <- glm(Length~Depth  + Year, data = nh,  family="gaussian")  
model3 <- glm(Length~Temp  + Year, data = nh,  family="gaussian")    
model4 <- glm(Length~Relief_score  + Year,data = nh, family="gaussian")
model5 <- glm(Length~Remoteness  + Year, data = nh,  family="gaussian") 
model6 <- glm(Length~Exposure  + Year, data = nh,  family="gaussian")
model7 <- glm(Length~Depth*Habitat  + Year, data = nh,  family="gaussian")   
model8 <- glm(Length~Temp*Habitat  + Year, data = nh,  family="gaussian")    
model9 <- glm(Length~Relief_score*Habitat  + Year,data = nh ,family="gaussian")
model10 <- glm(Length~Remoteness*Habitat  + Year, data = nh,  family="gaussian")  
model11 <- glm(Length~Exposure*Habitat  + Year, data = nh,  family="gaussian") 
model12 <- glm(Length~Relief_score + Depth  + Year, data = nh,  family="gaussian")    
model13 <- glm(Length~Relief_score + Remoteness  + Year, data = nh,  family="gaussian")  
model14 <- glm(Length~Relief_score + Exposure  + Year, data = nh,  family="gaussian") 
model15 <- glm(Length~Remoteness + Exposure  + Year, data = nh,  family="gaussian")  
model16 <- glm(Length~Depth + Remoteness  + Year,data = nh ,family="gaussian")

## model selection table
(msAICc <- model.sel(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
nh_df<-as.data.frame(msAICc,col_names = T)
nh_df$model <- row.names(nh_df)
nh_df$model <- as.numeric(substring(nh_df$model, 6))
nh_df$model<-as.character(nh_df$model)
nh_df$Species <- 'Nursehound'
nh_df <- nh_df %>% mutate(model = replace_na(model,'null_model'))

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(nh_df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs_nh <- left_join(model_outputs,pseudo_r_squared_df)

```


### 3.3.3 Pollack 
```{r}

## Pollack
pp <- subset(indicator_spp,indicator_spp$Species == 'Pollack')
pp <- pp %>%
  group_by(GRTS_ID) %>%
  summarise(Length = mean(Length))
pp <- merge(pp,envo_variables_short, by=c('GRTS_ID'))

hist(pp$Length)
mean_val_pp<-mean(pp$Length)
var(pp$Length)
std.error(pp$Length)
## model parameters
# Fit the null model (intercept only) using fitme
null_model <- glm(Length ~ 1, data = pp,  family="gaussian")
model1 <- glm(Length~Habitat + Year, data = pp,  family="gaussian") 
model2 <- glm(Length~Depth  + Year, data = pp,  family="gaussian")  
model3 <- glm(Length~Temp  + Year, data = pp,  family="gaussian")    
model4 <- glm(Length~Relief_score  + Year,data = pp, family="gaussian")
model5 <- glm(Length~Remoteness  + Year, data = pp,  family="gaussian") 
model6 <- glm(Length~Exposure  + Year, data = pp,  family="gaussian")
model7 <- glm(Length~Depth*Habitat  + Year, data = pp,  family="gaussian")   
model8 <- glm(Length~Temp*Habitat  + Year, data = pp,  family="gaussian")    
model9 <- glm(Length~Relief_score*Habitat  + Year,data = pp ,family="gaussian")
model10 <- glm(Length~Remoteness*Habitat  + Year, data = pp,  family="gaussian")  
model11 <- glm(Length~Exposure*Habitat  + Year, data = pp,  family="gaussian") 
model12 <- glm(Length~Relief_score + Depth  + Year, data = pp,  family="gaussian")    
model13 <- glm(Length~Relief_score + Remoteness  + Year, data = pp,  family="gaussian")  
model14 <- glm(Length~Relief_score + Exposure  + Year, data = pp,  family="gaussian") 
model15 <- glm(Length~Remoteness + Exposure  + Year, data = pp,  family="gaussian")  
model16 <- glm(Length~Depth + Remoteness  + Year,data = pp ,family="gaussian")

## model selection table
(msAICc <- model.sel(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
pp_df<-as.data.frame(msAICc,col_names = T)
pp_df$model <- row.names(pp_df)
pp_df$model <- as.numeric(substring(pp_df$model, 6))
pp_df$model<-as.character(pp_df$model)
pp_df$Species <- 'Pollock'
pp_df <- pp_df %>% mutate(model = replace_na(model,'null_model'))

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(pp_df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'
## merge
model_outputs_pp <- left_join(model_outputs,pseudo_r_squared_df)

## create dataset to predict to
modelx <- glm(Length~Depth + Remoteness  + Year,data = pp ,family="gaussian")
pdat <- expand.grid(Depth= seq(2.4,44,1),   Remoteness = mean(pp$Remoteness), Year=as.character(names(sort(table(pp$Year), decreasing = TRUE)[1])))
pred <- predict (modelx, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)

# Step 3: combine the predictions with the predictors, 
#         into a final dataframe (predframe)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se

# Add the predicted means
pp_p<-ggplot(predframe, aes(x=Depth, y=Length)) + 
  geom_line(aes(y = Length), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  geom_rug(aes(), data = pp,sides="b")+
  ylab('')+
  xlab('Depth (m)')+
  scale_x_continuous(limits = c(2.4,45), expand = c(0, 0)) +
  scale_y_continuous(limits = c(100,600),expand = c(0, 0)) +  
  coord_cartesian(xlim = c(min(predframe$Depth), max(predframe$Depth)), clip = "off")+
  theme_classic(base_size = 30)+
  theme(axis.title.y=element_blank())
pp_p

##remoteness
modely <-  glm(Length~Depth + Remoteness  + Year,data = pp ,family="gaussian") 
pdat <- expand.grid(Remoteness= seq(34,13200.3,1),   Depth = mean(pp$Depth), Year=as.character(names(sort(table(pp$Year), decreasing = TRUE)[1])))
pred <- predict (modely, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)
# Step 3: combine the predictions with the predictors, 
#         into a final dataframe (predframe)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se

# Add the predicted means
pp_p2<-ggplot(predframe, aes(x=Remoteness, y=Length)) + 
  geom_line(aes(y = Length), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  geom_rug(aes(), data = pp,sides="b")+
  ylab('Length (cm)')+
  xlab('Remoteness (m)')+
  scale_x_continuous(limits = c(0,13200), breaks = seq(0,13200,4000),expand = c(0, 0)) +
  scale_y_continuous(limits = c(250,600), expand = c(0, 0)) +
  coord_cartesian(xlim = c(min(predframe$Remoteness), max(predframe$Remoteness)), clip = "off")+
  theme_classic(base_size = 30)+
  theme(axis.title.y=element_blank())

pp_p2



```



### 3.3.4 Ballan wrasse
```{r}

## ballan wrasse
bw <- subset(indicator_spp,indicator_spp$Species == 'Ballan_wrasse')
bw <- bw %>%
  group_by(GRTS_ID) %>%
  summarise(Length = mean(Length))
bw <- merge(bw,envo_variables_short, by=c('GRTS_ID'))

hist(bw$Length)
mean(bw$Length)
var(bw$Length)
std.error(bw$Length)

## model parameters
# Fit the null model (intercept only) using fitme
null_model <- glm(Length ~ 1, data = bw,  family="gaussian")
model1 <- glm(Length~Habitat + Year, data = bw,  family="gaussian") 
model2 <- glm(Length~Depth + Year, data = bw,  family="gaussian")  
model3 <- glm(Length~Temp + Year, data = bw,  family="gaussian")    
model4 <- glm(Length~Relief_score + Year,data = bw, family="gaussian")
model5 <- glm(Length~Remoteness + Year, data = bw,  family="gaussian") 
model6 <- glm(Length~Exposure + Year, data = bw,  family="gaussian")
model7 <- glm(Length~Depth*Habitat + Year, data = bw,  family="gaussian")   
model8 <- glm(Length~Temp*Habitat + Year, data = bw,  family="gaussian")    
model9 <- glm(Length~Relief_score*Habitat + Year,data = bw ,family="gaussian")
model10 <- glm(Length~Remoteness*Habitat + Year, data = bw,  family="gaussian")  
model11 <- glm(Length~Exposure*Habitat + Year, data = bw,  family="gaussian") 
model12 <- glm(Length~Relief_score + Depth + Year, data = bw,  family="gaussian")    
model13 <- glm(Length~Relief_score + Remoteness + Year, data = bw,  family="gaussian")  
model14 <- glm(Length~Relief_score + Exposure + Year, data = bw,  family="gaussian") 
model15 <- glm(Length~Remoteness + Exposure + Year, data = bw,  family="gaussian")  
model16 <- glm(Length~Depth + Remoteness + Year,data = bw ,family="gaussian")

## model selection table
(msAICc <- model.sel(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
bw_df<-as.data.frame(msAICc,col_names = T)
bw_df$model <- row.names(bw_df)
bw_df$model <- as.numeric(substring(bw_df$model, 6))
bw_df$model<-as.character(bw_df$model)
bw_df$Species <- 'Ballan wrasse'
bw_df <- bw_df %>% mutate(model = replace_na(model,'null_model'))

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(bw_df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs_bw <- left_join(model_outputs,pseudo_r_squared_df)

## create dataset to predict to
## Remoteness
modelx <- glm(Length~Remoteness + Exposure + Year, data = bw,  family="gaussian") 
pdat <- expand.grid(Remoteness= seq(31.1,13185.7,1),   Exposure = mean(bw$Exposure), Year=as.character(names(sort(table(bw$Year), decreasing = TRUE)[1])))
pred <- predict (modelx, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)

# Step 3: combine the predictions with the predictors, 
#         into a final dataframe (predframe)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se

# Add the predicted means
bw_p<-ggplot(predframe, aes(x=Remoteness, y=Length)) + 
  geom_line(aes(y = Length), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  geom_rug(aes(), data = bw,sides="b")+
  ylab('')+
  xlab('Remoteness (m)')+
  scale_x_continuous(limits = c(0,13200), breaks = seq(0,13200,4000),expand = c(0, 0)) +
  scale_y_continuous(limits = c(250,450),expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(axis.title.y=element_blank())
  
bw_p

## create dataset to predict to
modely <- glm(Length~Remoteness + Exposure + Year, data = bw,  family="gaussian") 
pdat <- expand.grid(Exposure= seq(0.8,173.6,1),   Remoteness = mean(bw$Remoteness), Year=as.character(names(sort(table(bw$Year), decreasing = TRUE)[1])))
pred <- predict (modely, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)

# Step 3: combine the predictions with the predictors, 
#         into a final dataframe (predframe)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se

# Add the predicted means
bw_p2<-ggplot(predframe, aes(x=Exposure, y=Length)) + 
  geom_line(aes(y = Length), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  geom_rug(aes(), data = bw,sides="b")+
  ylab('')+
  xlab('Exposure ()')+
  scale_x_continuous(limits = c(0,180), expand = c(0, 0)) +
  scale_y_continuous(limits = c(220,450),expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(axis.title.y=element_blank())


bw_p2


```

### 3.3.5 Cuckoo wrasse
```{r}

## Cuckoo wrasse
cw <- subset(indicator_spp,indicator_spp$Species == 'Cuckoo_wrasse')
cw <- cw %>%
  group_by(GRTS_ID) %>%
  summarise(Length = mean(Length))
cw <- merge(cw,envo_variables_short, by=c('GRTS_ID'))

hist(cw$Length)
mean(cw$Length)
var(cw$Length)
std.error(cw$Length)

## model parameters
# Fit the null model (intercept only) using fitme
null_model <- glm(Length ~ 1, data = cw,  family="gaussian")
model1 <- glm(Length~Habitat + Year, data = cw,  family="gaussian") 
model2 <- glm(Length~Depth + Year, data = cw,  family="gaussian")  
model3 <- glm(Length~Temp + Year, data = cw,  family="gaussian")    
model4 <- glm(Length~Relief_score + Year,data = cw, family="gaussian")
model5 <- glm(Length~Remoteness + Year, data = cw,  family="gaussian") 
model6 <- glm(Length~Exposure + Year, data = cw,  family="gaussian")
model7 <- glm(Length~Depth*Habitat + Year, data = cw,  family="gaussian")   
model8 <- glm(Length~Temp*Habitat + Year, data = cw,  family="gaussian")    
model9 <- glm(Length~Relief_score*Habitat + Year,data = cw ,family="gaussian")
model10 <- glm(Length~Remoteness*Habitat + Year, data = cw,  family="gaussian")  
model11 <- glm(Length~Exposure*Habitat + Year, data = cw,  family="gaussian") 
model12 <- glm(Length~Relief_score + Depth + Year, data = cw,  family="gaussian")    
model13 <- glm(Length~Relief_score + Remoteness + Year, data = cw,  family="gaussian")  
model14 <- glm(Length~Relief_score + Exposure + Year, data = cw,  family="gaussian") 
model15 <- glm(Length~Remoteness + Exposure + Year, data = cw,  family="gaussian")  
model16 <- glm(Length~Depth + Remoteness + Year,data = cw ,family="gaussian")

## model selection table
(msAICc <- model.sel(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
cw_df<-as.data.frame(msAICc,col_names = T)
cw_df$model <- row.names(cw_df)
cw_df$model <- as.numeric(substring(cw_df$model, 6))
cw_df$model<-as.character(cw_df$model)
cw_df$Species <- 'Cuckcoo wrasse'
cw_df <- cw_df %>% mutate(model = replace_na(model,'null_model'))

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(cw_df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs_cw <- left_join(model_outputs,pseudo_r_squared_df)

## plot the top variables present in the optimal model
## generate modeled values for depth and relief against richness
## depth
## for reef
modelx <- glm(Length~Depth*Habitat + Year, data = cw,  family="gaussian")   
pdat <- expand.grid(Depth = seq(18.2,44.2,1),  Habitat = "Reefs", Year=as.character(names(sort(table(cw$Year), decreasing = TRUE)[1])))
pred <- predict (modelx, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se

## for sediment
pdat2 <- expand.grid(Depth = seq(4.4,44,1),  Habitat = "Sediment", Year=as.character(names(sort(table(cw$Year), decreasing = TRUE)[1])))
pred2 <- predict (modelx, newdata = pdat2, na.rm = T, type= "response", se.fit = TRUE)
predframe2 <- data.frame (pdat2, Length = pred2$fit, se = pred2$se.fit)
predframe2$lwr <- predframe2$Length - 1.96*predframe2$se
predframe2$upr <- predframe2$Length + 1.96*predframe2$se
## bind
predframe3 <- rbind(predframe,predframe2)

cw_p <- ggplot(mapping = aes(x = Depth)) +
  geom_line(aes(y = Length), data = predframe,linewidth=1, colour = "#077DAA") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2, fill = "#077DAA")+
  geom_rug(aes(), data = cw)+
  labs(x = 'Depth (m)')+
  labs(y = '')+
  scale_x_continuous(limits = c(18,45), expand = c(0, 0)) +
  scale_y_continuous(limits = c(200,300), expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  #theme(legend.position = "inside", legend.position.inside = c(.15, .75),legend.text=element_text(size=18))
  theme(legend.position = "none",)+
  theme(axis.title.y=element_blank())
cw_p 


```


### 3.3.1 Brown crab
```{r}

## Brown crab
ec <- subset(indicator_spp,indicator_spp$Species == 'Edible_crab')
ec <- ec %>%
  group_by(GRTS_ID) %>%
  summarise(Length = mean(Length))
ec <- merge(ec,envo_variables_short, by=c('GRTS_ID'))

hist(ec$Length)
mean(ec$Length)
var(ec$Length)
std.error(ec$Length)

## model parameters
# Fit the null model (intercept only) using fitme
null_model <- glm(Length ~ 1, data = ec,  family="gaussian")
model1 <- glm(Length~Habitat +Year, data = ec,  family="gaussian") 
model2 <- glm(Length~Depth +Year, data = ec,  family="gaussian")  
model3 <- glm(Length~Temp +Year, data = ec,  family="gaussian")    
model4 <- glm(Length~Relief_score +Year,data = ec, family="gaussian")
model5 <- glm(Length~Remoteness +Year, data = ec,  family="gaussian") 
model6 <- glm(Length~Exposure +Year, data = ec,  family="gaussian")
model7 <- glm(Length~Depth*Habitat +Year, data = ec,  family="gaussian")   
model8 <- glm(Length~Temp*Habitat +Year, data = ec,  family="gaussian")    
model9 <- glm(Length~Relief_score*Habitat +Year,data = ec ,family="gaussian")
model10 <- glm(Length~Remoteness*Habitat +Year, data = ec,  family="gaussian")  
model11 <- glm(Length~Exposure*Habitat +Year, data = ec,  family="gaussian") 
model12 <- glm(Length~Relief_score + Depth +Year, data = ec,  family="gaussian")    
model13 <- glm(Length~Relief_score + Remoteness +Year, data = ec,  family="gaussian")  
model14 <- glm(Length~Relief_score + Exposure +Year, data = ec,  family="gaussian") 
model15 <- glm(Length~Remoteness + Exposure +Year, data = ec,  family="gaussian")  
model16 <- glm(Length~Depth + Remoteness +Year,data = ec ,family="gaussian")

## model selection table
(msAICc <- model.sel(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
ec_df<-as.data.frame(msAICc,col_names = T)
ec_df$model <- row.names(ec_df)
ec_df$model <- as.numeric(substring(ec_df$model, 6))
ec_df$model<-as.character(ec_df$model)
ec_df$Species <- 'Edible crab'
ec_df <- ec_df %>% mutate(model = replace_na(model,'null_model'))

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(ec_df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs_ec <- left_join(model_outputs,pseudo_r_squared_df)

## create dataset to predict to
##  Temp
modelx <- glm(Length~Temp +Year, data = ec,  family="gaussian")
pdat <- expand.grid(Temp = seq(11.0,17,0.1), Year=as.character(names(sort(table(ec$Year), decreasing = TRUE)[1])))
pred <- predict (modelx, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)

# Step 3: combine the predictions with the predictors, 
#         into a final dataframe (predframe)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se

# Add the predicted means
ec_p<-ggplot(predframe, aes(x=Temp, y=Length)) + 
  geom_line(aes(y = Length), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  geom_rug(aes(), data = ec,sides="b")+
  ylab('')+
  xlab('Temp (c)')+
  scale_x_continuous(limits = c(10.9,17), expand = c(0, 0)) +
  scale_y_continuous(limits = c(100,175),expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(axis.title.y=element_blank())
  
ec_p



```


### 3.3.1 Spiny lobster
```{r}

## Common_spiny_lobster
sl <- subset(indicator_spp,indicator_spp$Species == 'Common_spiny_lobster')
mean(sl$Length)
var(sl$Length)
sd(sl$Length)
max(sl$Length)

sl_lgnth <- sl[order(sl$Length),]

sl <- sl %>%
  group_by(GRTS_ID) %>%
  summarise(Length = mean(Length))
sl <- merge(sl,envo_variables_short, by=c('GRTS_ID'))

hist(sl$Length)
mean(sl$Length)
var(sl$Length)
sd(sl$Length)
min(sl$Length)

## model parameters
# Fit the null model (intercept only) using fitme
null_model <- glm(Length ~ 1, data = sl,  family="gaussian")
model1 <- glm(Length~Habitat + Year, data = sl,  family="gaussian") 
model2 <- glm(Length~Depth + Year, data = sl,  family="gaussian")  
model3 <- glm(Length~Temp + Year, data = sl,  family="gaussian")    
model4 <- glm(Length~Relief_score + Year,data = sl, family="gaussian")
model5 <- glm(Length~Remoteness + Year, data = sl,  family="gaussian") 
model6 <- glm(Length~Exposure + Year, data = sl,  family="gaussian")
model7 <- glm(Length~Depth*Habitat + Year, data = sl,  family="gaussian")   
model8 <- glm(Length~Temp*Habitat + Year, data = sl,  family="gaussian")    
model9 <- glm(Length~Relief_score*Habitat + Year,data = sl ,family="gaussian")
model10 <- glm(Length~Remoteness*Habitat + Year, data = sl,  family="gaussian")  
model11 <- glm(Length~Exposure*Habitat + Year, data = sl,  family="gaussian") 
model12 <- glm(Length~Relief_score + Depth + Year, data = sl,  family="gaussian")    
model13 <- glm(Length~Relief_score + Remoteness + Year, data = sl,  family="gaussian")  
model14 <- glm(Length~Relief_score + Exposure + Year, data = sl,  family="gaussian") 
model15 <- glm(Length~Remoteness + Exposure + Year, data = sl,  family="gaussian")  
model16 <- glm(Length~Depth + Remoteness + Year,data = sl ,family="gaussian")

## model selection table
(msAICc <- model.sel(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16))
sl_df<-as.data.frame(msAICc,col_names = T)
sl_df$model <- row.names(sl_df)
sl_df$model <- as.numeric(substring(sl_df$model, 6))
sl_df$model<-as.character(sl_df$model)
sl_df$Species <- 'Common spiny lobster'
sl_df <- sl_df %>% mutate(model = replace_na(model,'null_model'))

## extract the Coefficients and Standard Errors
model_list<-list(null_model,model1,model2,model3,model4,model5,model6,model7,model8,model9,model10,model11,model12,model13,model14,model15,model16) 
model_summaries <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  coef_data <- as.data.frame(coefTable(model_list[[i]]))
  coef_data <- coef_data[rownames(coef_data) == "(Intercept)", ]
  coef_data$model <- paste0(i-1)
  coef_data
}))
model_summaries <- within(model_summaries, {
  CI_lower <- Estimate - 1.96 * `Std. Error`
  CI_upper <- Estimate + 1.96 * `Std. Error`
})

model_summaries[1,4]<- 'null_model'
model_summaries<-model_summaries[,c(4,5,6)]

## merge
model_outputs <- left_join(sl_df,model_summaries)
  
# Calculate the pseudo R-squared 
pseudo_r_squared_df <- do.call(rbind, lapply(seq_along(model_list), function(i) {
  model <- model_list[[i]]
  pseudo_r_squared <- (model$null.deviance - model$deviance) / model$null.deviance
  data.frame(
    model = paste0(i-1),
    pseudo_r_squared = pseudo_r_squared
  )
}))

# Reset row names for clarity
rownames(pseudo_r_squared_df) <- NULL
pseudo_r_squared_df[1,1]<- 'null_model'

## merge
model_outputs_sp <- left_join(model_outputs,pseudo_r_squared_df)

## create dataset to predict to
## Remoteness
modelx <- glm(Length~Depth , data = sl,  family="gaussian")
pdat <- expand.grid(Depth= seq(19,40,0.5), Year=as.character(names(sort(table(sl$Year), decreasing = TRUE)[1])))
pred <- predict (modelx, newdata = pdat, na.rm = T, type= "response", se.fit = TRUE)

# Step 3: combine the predictions with the predictors, 
#         into a final dataframe (predframe)
predframe <- data.frame (pdat, Length = pred$fit, se = pred$se.fit)
predframe$lwr <- predframe$Length - 1.96*predframe$se
predframe$upr <- predframe$Length + 1.96*predframe$se

# Add the predicted means
sl_p<-ggplot(predframe, aes(x=Depth, y=Length)) + 
  geom_line(aes(y = Length), data = predframe,linewidth=1,col='darkblue') +
  geom_ribbon(aes(ymin = lwr, ymax = upr), data = predframe, alpha = 0.2)+
  geom_rug(aes(), data = sl,sides="b")+
  ylab('Length (cm)')+
  xlab('Depth (m)')+
  scale_x_continuous(limits = c(19,40), expand = c(0, 0)) +
  scale_y_continuous(limits = c(100,160),expand = c(0, 0)) +
  theme_classic(base_size = 30)+
  theme(axis.title.y=element_blank())

sl_p

```

#### 4.1 Export lengths results for Figure 6
```{r}

## merge and export all model results for supp. table 
length_results<- rbind(model_outputs_ssc,model_outputs_nh,model_outputs_pp,model_outputs_bw,model_outputs_cw,model_outputs_ec,model_outputs_sp)

species_lengths_models <-  grid.arrange(ssc_p,pp_p,pp_p2,bw_p,bw_p2,cw_p,ec_p,sl_p,ncol =3)


```

